Three Problems for Sophistication

### 2015-01-19 Facticity/Facticity/paper.tex
nog wat verder gelezen

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $S^\T_K(x) \geq K(x)/6$ and $S^\T(x) \geq K(x)/6$. 
\end{restatable} 
\noindent This tells us that $S^\T_K(x)$ and $S^\T(x)$ can truly grow to a significant part of the total description length. While this shows that sophistication, if defined with particular care, is not entirely vacuous, we can shed some light on exactly which strings end up with high sophistication, using the principle of \emph{depth}. We will use the following definition:

@@

\begin{restatable}{theorem}{depth}
Let $A(n)$ be the single-argument Ackermann function and $c$ some arbitrary constant. There are numberings such that for all strings with depth $d^\C_c(x) \leq A(C(x))$, $\s^\T(x)$, $s^\P(x)$ and $\s^\F$ are bounded.
\end{restatable}
\noindent This shows that while high-sophistication strings exist, the mechanics do not follow the basic intuition of sophistication. Unless we encounter data that would take longer than the lifetime of the universe (and then some) to produce, sophistication behaves exactly as it would without the restricted model class: a solution to a super-exponential problem would have no greater sophistication than a string sampled from a simple Bernoulli model. This result also suggests that the `non-stochastic' property of strings with high sophistication\cite{shen1983concept,vereshchagin2004kolmogorov} says more about depth and totality than it does aboutrandomness and structure.

@@

\section{Conclusion}
\label{section:conclusion}
As we have seen, the idea behind sophistication has a rich history. It has been proposed many times, by many different authors. This tells us two things. Firstly, there is a strong intuition behind sophistication, shared by many respected authors, and second, it is not easy to get the definition exactly right.

So why should we consider this intuition reasonable at all? The first assumption of sophistication is that high compressibility and pure randomness are objectively `uninteresting'. A more sophisticated signal, such as a television broadcast must contain, in some objective sense, more meaningful information. Whether this intuition seems reasonable likely differs per person, but sophistication goes one step further: it also suggests that we can objectively separate the information in a string into random and structured.

To see whether this intuition is valid; imagine that we sample from the universal distribution by feeding random bits to a universal Turing machine. If this UTM is constructed in the conventional manner, the initial bits will determine the index of a partial computable function $i$, and the rest will be the input to that machine $y$. What's the characteristic model of this data? Certainly, we cannot exclude the possibility that $\phi_i$ is also a universal model. Also, we cannot exclude the possibility that $\phi_i$ is a singleton model for $x$.

Although sampling perspectives are discouraged by proponents of algorithmic statistics, this example does clash with the intuition behind sophistication: there is no reason to assume that a string with high sophistication was sampled from a complex model, so how can a complex model be intrinsic to the data? \footnotemark

@@

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. The candidate set for this string contains various models from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal models says that it is a `thing'. A more specific model might say that it is an image. Even more specific would be a painting, a Monet, or specifically the painting Impression of a sunrise. These are the models we imagine we may find in the candidate set. Can we say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? We see no intuitive reason why such a distinction should be possible.

Of course, Kolmogorov complexity allows us to say that it is unlikely to be a Jackson Pollock, or a piece of music. And, if we were to get a second sample from the same source, we \emph{could} answer the question: we would receive either the same painting again, or perhaps another Monet. But, a single sample does not give us enough evidence to choose a single model. Ultimately, this is just a metaphor, but since the notion of sophistication is driven by intuition, we feel that an intuition to the contrary should serve to sharpen the debate.

For this reason, we take a skeptical view of sophistication. We can offer the following insight, to back up our skepticism. As we have seen, candidates within a constant of the optimal representation can occur all over the spectrum, from small to large models. A small change in construction, such as a different acceptable numbering, or a change in threshold function, can arbitrarily push models into or out of the candidate set. Whereas in Kolmogorov complexity, constant changes have constant effects on the outcome, here constant changes can cause large changes in the outcome. While no authors have touched on the issue of invariance, it is related to the issue of instability \cite{antunes2013sophistication,vereshchagin2013algorithmic}. We consider invariance the first criterion by which any proposal must be judged, with the burden of proof resting with the proposer, rather than the critic.

Finally, we note that the use of total functions has a very subtle effect on the discussion surrounding sophistication. Since the total functions are not enumerable, the value of the sophistication relies strongly on the difference between the Kolmogorov complexity and the Kolmogorov complexity bounded to total functions. This gap is exceedingly difficult to capture, because the total functions cannot be enumerated. This is dangerous: if such definitions cannot be proved wrong or right, they will stand as the ad-hoc official view, while the restrictions used to elude proofs of incorrectness make them very difficult to build on. This creates an artificial dead end for a very valuable area of research.

>>>>

\begin{restatable}{theorem}{dogfood}
There are infinitely many $x$ with $S^\T_K(x) \geq K(x)/6$ and $S^\T(x) \geq K(x)/6$. 
\end{restatable} 
\noindent This tells us that $S^\T_K(x)$ and $S^\T(x)$ can truly grow to a significant part of the total description length. While this shows that sophistication, if defined with particular care, is not entirely vacuous, \sdr{hold on, it might still be completely meaningless and/or not invariant} we can shed some light on exactly which strings end up with high sophistication, using the principle of \emph{depth}. We will use the following definition:

@@

\begin{restatable}{theorem}{depth}
Let $A(n)$ be the single-argument Ackermann function and $c$ some arbitrary constant. There are numberings such that for all strings with depth $d^\C_c(x) \leq A(C(x))$, $\s^\T(x)$, $s^\P(x)$ and $\s^\F$ are bounded.
\end{restatable}
\noindent This shows that while high-sophistication strings exist, the mechanics do not follow the basic intuition of sophistication. Unless we encounter data that would take longer than the lifetime of the universe (and then some) to produce \sdr{Not sure about this philosophically, but lets discuss}, sophistication behaves exactly as it would without the restricted model class: a solution to a super-exponential problem \sdr{Bad example, use something that clearly does not require much computation but has a lot of structure. A midpoint-displaced heightmap comes to mind} would have no greater sophistication than a string sampled from a simple Bernoulli model. This result also suggests that the `non-stochastic' property of strings with high sophistication\cite{shen1983concept,vereshchagin2004kolmogorov} says more about depth and totality than it does about structure and noise.

@@

\section{Discussion and Conclusion}
\label{section:conclusion}
As we have seen, the idea behind sophistication has a rich history. It has been proposed many times, by many different authors. This tells us two things. First, there is a strong intuition behind sophistication, shared by many respected authors, and second, it is not easy to get the definition exactly right.

So why should we consider this intuition reasonable at all? The first assumption of sophistication is that high compressibility and pure randomness are objectively `uninteresting'. A more sophisticated signal, such as a television broadcast must contain, in some objective sense, more meaningful information. Whether this intuition seems reasonable likely differs per person \sdr{weak}, but sophistication goes one step further: it also suggests that we can objectively separate the information in a string into random and structured. \sdr{I think here we should sell the intuition better before we tear it down.}

To see whether this intuition is valid; imagine that we sample from the universal distribution by feeding random bits to a universal Turing machine, machine $u$ in the $\phi$-enumeration. If this UTM is constructed in the conventional manner, the initial bits will determine the index of a partial computable function $i$, and the rest will be the input $y$ to that machine. What is the characteristic model of the resulting data? Certainly, if we have to judge based only on the data we cannot exclude the possibility that the data was sampled from $\phi_u$: after all, it was.  On the other hand, neither can we exclude the possibility that it was generated by $\phi_i$, as again, it was!

Although sampling perspectives are discouraged by proponents of algorithmic statistics, this statistical sanity check illustrates why it seems unjustified to assume that, based on only the data, all but one model can always be disqualified. \sdr{Hier stond: There is no reason to assume that a string with high sophistication was sampled from a complex model, so how can a complex model be intrinsic to the data? \footnotemark}

@@

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. The candidate set for this string contains various models from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal models says that it is `something computable'. A more specific model might say that it is `an image'. Even more specific would be a painting, a Monet, or specifically the painting Impression of a sunrise. These are the models we imagine we may find in the candidate set. Can we say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? We see no intuitive reason why such a distinction should be possible without making any further assumptions.

Of course, \sdr{Actually this is not trivial at all} Kolmogorov complexity allows us to say that it is unlikely to be a Jackson Pollock, or a piece of music. And, if we were to get a second sample from the same source, we \emph{could} answer the question: we would receive either the same painting again, or perhaps another Monet. But, a single sample does not give us enough evidence to choose a single model. \sdr{Volgende zin weer zwak, de metafoor wordt backed up door ons voorbeeld hierboven dus is behoorlijk hard} Ultimately, this is just a metaphor, but since the notion of sophistication is driven by intuition, we feel that an intuition to the contrary should serve to sharpen the debate.

For this reason, we take a skeptical view of sophistication. We can offer the following insight, to back up our skepticism. As we have seen, candidates within a constant of the optimal representation can occur all over the spectrum, from small to large models. A small change in construction, such as a different acceptable numbering, or a change in threshold function, can arbitrarily push models into or out of the candidate set. Whereas in Kolmogorov complexity, constant changes have constant effects on the outcome, here constant changes can cause large changes in the outcome. \sdr{Not sufficiently clear} While no authors have touched on the issue of invariance, it is related to the issue of instability \cite{antunes2013sophistication,vereshchagin2013algorithmic}. We consider invariance the first criterion by which any proposal must be judged, with the burden of proof resting with the proposer, rather than the critic.

Finally, we note that the use of total functions has a very subtle effect on the discussion surrounding sophistication. Since the total functions are not enumerable, the value of the sophistication relies strongly on the difference between the Kolmogorov complexity and the Kolmogorov complexity restricted to total functions. This gap is exceedingly difficult to capture, because the total functions cannot be enumerated. This is dangerous: if such definitions cannot be proved wrong or right, they will stand as the ad-hoc official view, while the restrictions used to elude proofs of incorrectness make them very difficult to build on. This creates an artificial dead end for a very valuable area of research.

\sdr{Something about what is useful about the theory}
\sdr{A picture of the structure function (perhaps sooner as well)}
\sdr{This has become a discussion section. It does not recap. That's a bit unusual.}
